### 2016-04-10 text/Chapter08.tex
Implementation chapter

\item We consider a number of case studies (Section~\ref{sec:impl-case}) that illustrate ointeresting
  aspects of the theories discussed earlier. This includes the typing of lambda abstraction and the
  difference between flat and structural systems (Section~\ref{sec:impl-case-typing}) and the
  comonadically-inspired translation (Section~\ref{sec:impl-case-transl}).

@@

\subsection{Type checking}
\label{sec:impl-theory-typing}

(Section~\ref{sec:flat-unique})

\paragraph{Implementation.}
From the presentation in this section, it might appear that resolving the ambiguity related to
lambda abstraction for implicit parameters requires a type system that is quite different from
the core flat coeffect type system shown earlier in Figure~\ref{fig:flat-types}. This is not the
case. As discussed in Chapter~\ref{ch:impl}, the required changes in the implementation are
simpler.

Briefly, the implementation collects constraints on the coeffects and then finds the smallest sets
of implicit parameters to satisfy the constraints. We still need to track implicit parameters in
scope $\cclrd{\Delta}$, but the rest of the (\emph{abs}) rule from the implementation is close
to the one from Figure~\ref{fig:flat-types}:
%
\begin{equation*}
\tyrule{abs}
  {\coctx{\Gamma, x\!:\!\tau_1;\cclrd{\Delta}}{\cclrd{t}} \vdash e : \tau_2 ~|~ C}
  {\coctx{\Gamma;\cclrd{\Delta}}{\cclrd{r}} \vdash \lambda x\!:\!\tau_1.e : \tau_1 \xrightarrow{\cclrd{s}} \tau_2 ~|~
    C\cup\{\cclrd{t}=\cclrd{r} \,\czip\, \cclrd{s}, \cclrd{r}=\cclrd{\Delta} \}}
\end{equation*}
%
Given a typing derivation for the body that produced constraints $C$, we generate an additional
constraint that restricts $\cclrd{r}$ (declaration-site demands) to those available in the
current static scope $\cclrd{\Delta}$. The constraint satisfaction algorithm then finds
the minimal set $\cclrd{s}$ which is $\cclrd{t}\cclrd{\setminus \Delta}$.

\paragraph{Implementation.}
As with implicit parameters, the implementation (discussed in Chapter~\ref{ch:impl}) does not
require changing the typing (\emph{abs}) rule of the flat coeffect system. The typing specified
by (\emph{idabs}) can be easily obtained by generating additional constraints:
%
\begin{equation*}
\tyrule{abs}
  {\coctx{\Gamma, x\!:\!\tau_1}{\cclrd{t}} \vdash e : \tau_2 ~|~ C}
  {\coctx{\Gamma}{\cclrd{s}} \vdash \lambda x\!:\!\tau_1.e : \tau_1 \xrightarrow{\cclrd{t}} \tau_2 ~|~
    C\cup\{\cclrd{t}=\cclrd{r} \,\czip\, \cclrd{s}, \cclrd{r}=\cclrd{t}, \cclrd{s}=\cclrd{t} \}}
\end{equation*}
%
Here, the two additional constraints restrict both $\cclrd{r}$ and $\cclrd{s}$ to be equal to the
coeffect of the body $\cclrd{t}$ and so the only possible resolution is the one specified by
(\emph{idabs}).

Type checking
 * DSL specific constraint resolution
 * easier for structural

\subsection{Translation}
\label{sec:impl-theory-transl}

Translation ~ Semantics

\subsection{Extensions}
\label{sec:impl-theory-ext}

It is really easy to add another DSL

@@

\subsection{Typing}
\label{sec:impl-case-typing}

let both =
  let ?fst = 100 in
  fun trd -> ?fst + ?snd + trd in
let ?fst = 200 in
both 1

num -{ ?snd:num }-> num

let struct x y = x + prev y in
struct

num -{ 1 }-> num -{ 1 }-> num

let flat x y = x + prev y in
flat

num -{ 0 }-> num -{ 1 }-> num

@@

let ?param = 10 in
fun x -> ?param + ?other

let (ctx2, ctx3) = split (duplicate finput) in
let ctx1 = letimpl ?param (ctx2, 10) in
fun x ->
  let ctx4 = merge (x, ctx1) in
  let (ctx5, ctx6) = split (duplicate ctx4) in
  lookup ?param ctx5 + lookup ?other ctx6

fun x -> (fun v -> prev v) (prev x)

fun x ->
  let ctx1 = merge (x, sinput) in
  (fun v ->
    let ctx3 = merge (v, choose_0 ctx1) in
    counit (prev ctx3)) (cobind (fun ctx2 ->
      counit (prev ctx2)) (choose_1 ctx1))

>>>>

\item We consider a number of case studies (Section~\ref{sec:impl-case}) that illustrate interesting
  aspects of the theories discussed earlier. This includes the typing of lambda abstraction and the
  difference between flat and structural systems (Section~\ref{sec:impl-case-typing}) and the
  comonadically-inspired translation (Section~\ref{sec:impl-case-transl}).

@@

% --------------------------------------------------------------------------------------------------

\subsection{Type checking and inference}
\label{sec:impl-theory-typing}

To simplify the writing of context-aware programs, the implementation provides a limited form of
type inference (Section~\ref{sec:impl-tech}). This is available just for convenience and so we do
not claim any completeness or complexity result about the algorithm and we do not present full
formalization. However, it is worth noting how the domain-specific procedures for choosing a unique
type derivation (Section~\ref{sec:flat-unique}) are adapted.

The type inference works in the standard way \cite{types-mlessence,types-inference} by generating
type constraints and solving them. Solving of type constraints is done in the standard way, but we
additionally collect and solve \emph{coeffect constraints}. In order to obtain unique type
derivation, we generate additional coeffect constraints in lambda abstraction of each flat
coeffect language.

\paragraph{Flat implicit parameters.}

As discussed in Section~\ref{sec:flat-unique}, when choosing unique typing derivation for
implicit parameters, we keep track of the implicit parameters available in lexical scope
(written as $\cclrd{\Delta}$). In lambda abstraction rule, the implicit parameters required by
the body (tracked by $\ccclrd{r}$) are split so that all parameters available in lexical scope
are captured and only the remaining parameters ($\cclrd{r}\setminus\cclrd{\Delta}$) are required
from the caller of the function.

From the presentation in Section~\ref{sec:flat-unique}, it might appear that resolving the ambiguity
related to lambda abstraction for implicit parameters requires a type system different from
the core flat coeffect type system shown earlier in Section ~\ref{sec:flat-calculus-types}.
This is not the case.

We track implicit parameters in scope $\cclrd{\Delta}$, but the rest of the (\emph{abs}) rule from
the implementation only generates an additional coeffect constraint. Writing
$\coctx{\Gamma}{\cclrd{t}} \vdash e : \tau ~|~ C}$ for a judgement that generates coeffect
constraints $C$, the (\emph{abs}) rule used for implicit parameters looks as follows:
%
\begin{equation*}
\tyrule{abs}
  {\coctx{\Gamma, x\!:\!\tau_1;\cclrd{\Delta}}{\cclrd{t}} \vdash e : \tau_2 ~|~ C}
  {\coctx{\Gamma;\cclrd{\Delta}}{\cclrd{r}} \vdash \lambda x\!:\!\tau_1.e : \tau_1 \xrightarrow{\cclrd{s}} \tau_2 ~|~
    C\cup\{\cclrd{t}=\cclrd{r} \,\czip\, \cclrd{s}, \cclrd{r}=\cclrd{\Delta} \}}
\end{equation*}
%
Given a typing derivation for the body that produced constraints $C$, we generate an additional
constraint that restricts $\cclrd{r}$ (declaration-site demands) to those available in the
current static scope $\cclrd{\Delta}$. It is not necessary to generate a constraint for the
coeffect $\cclrd{s}$, because our constraint satisfaction algorithm finds
the minimal set $\cclrd{s}$ which is $\cclrd{t}\cclrd{\setminus \Delta}$.

\paragraph{Flat data-flow.}
In context-aware language for data-flow (and in language with liveness tracking), the inherent
ambiguity of the (\emph{abs}) rule is resolved by duplicating the context requirements of the body.
In Section~\ref{sec:flat-unique}, this was defined by replacing the standard coeffect (\emph{abs})
rule with a rule (\emph{idabs}) that uses an annotation $\cclrd{r}$ for the body of the function,
declaration-site coeffect and call-site coeffect.

As with implicit parameters, the implementation does not require changing the core (\emph{abs})
typing rule of the flat coeffect system. Instead, the unique resulution is obtained by generating
additional coeffect constraints:
%
\begin{equation*}
\tyrule{abs}
  {\coctx{\Gamma, x\!:\!\tau_1}{\cclrd{t}} \vdash e : \tau_2 ~|~ C}
  {\coctx{\Gamma}{\cclrd{s}} \vdash \lambda x\!:\!\tau_1.e : \tau_1 \xrightarrow{\cclrd{t}} \tau_2 ~|~
    C\cup\{\cclrd{t}=\cclrd{r} \,\czip\, \cclrd{s}, \cclrd{r}=\cclrd{t}, \cclrd{s}=\cclrd{t} \}}
\end{equation*}
%
Here, the two additional constraints restrict both $\cclrd{r}$ and $\cclrd{s}$ to be equal to the
coeffect of the body $\cclrd{t}$ and so the only possible resolution is the one specified by
(\emph{idabs}).

% --------------------------------------------------------------------------------------------------

\subsection{Execution of context-aware programs}
\label{sec:impl-theory-transl}

Context-aware programs are executed by translating the source program into a simple functional
target language. For simplicity, programs in the simple target language are then interpreted, but
they could equally be compiled using standard techniques for compiling functional code. The
translation follows the rules defined in Section~\ref{sec:semantics-translation} (for flat coeffect
languages) and Section~X (for structural coeffect languages). The result of the translation is a
program that consists of the following:

\begin{itemize}
\item {\sc Functional constructs.} Those include binary operations, tuples, let binding,
  constants, variables, function abstraction and application. The interpreter keeps a map of
  assignments for variables in scope and recursively evaluates the expression.

\item {\sc Comonadic operations.} Those are the comonadic primitives provided by indexed comonads --
  \ident{cobind}, \ident{counit} together with \ident{merge} and \ident{split} for flat coeffects
  or \ident{merge} and \ident{choose} for structural coeffects. The translation that produces these
  is shared by all context-aware languages, but their definition in the interpreter is
  domain-specific.

\item {\sc Domain-specific operations.} Each context-aware language may additionally include
  operations that model domain-specific operations. For data-flow, this is \ident{prev} (accessing
  past values) and for implicit parameters, this is \ident{letimpl} and \ident{lookup} for implicit
  parameter binding and access, respectively.
\end{itemize}

\noindent
The fact that the prototype implementation is based on the theoretical framework provided by
coeffect calculi means that it has the desirable properties proved in Section~\ref{sec:semantics-proofs}
and Section~X. In particular, evaluating a well-typed context-aware program in a context that
provides sufficient contextual capabilities will not cause an error.

In the interactive essay (Section~\ref{sec:impl-essay}), we further use the coeffects to
automatically generate a user interface that requires the user to provide the required contextual
capabilities (past values for individual variables, or values for implicit parameters).

Another benefit of using the common framework is that the implementation can be easily extended
to support additional context-aware languages.

% --------------------------------------------------------------------------------------------------

\subsection{Supporting more context-aware languages}
\label{sec:impl-theory-ext}

The prototype implementation supports two of the context-aware languages discussed in this thesis:
implicit parameters and data-flow. The remaining examples are calculus for tracking variable
liveness (of ordinary variables) and the structural language based on bounded reuse (counting
number of times values are used). However, the prototype implementation is based on the
common coeffect framework and makes it easy to add support for these and also for other context
aware languages based on coeffects.

In order to extend the implementation with support for liveness or bounded reuse tracking (or
other context-aware language), the following 4 additions are required:

\begin{enumerate}
\item A domain-specific function abstraction rule that resolves the ambiguity in the general
  (\emph{abs}) rule of the flat coeffect calculus. For liveness, the handling would be the same
  as for data-flow, but for other flat coeffect systems, another resolution mechanism might be used
  instead.

\item A domain-specific instance of coeffect algebra needs to be provided. In order to support
  the type inference in the prototype implementation, the constraint solver needs to be extended
  to solve constraint using the coeffect algebra. For liveness, this would be solving simple
  two-point lattice constraints.

\item For evaluation, a new kind of comonadic values needs to be added. For liveness, this would be
  an option value that may or may not contain a value. The semantics of comonadic operations on the
  values needs to be defined.

\item For context-aware languages that have additional primitives (such as \kvd{prev} or \ident{?param}),
  the parser and AST needs to be extended, custom type-checking and translation rules added and
  domain-specific primitive operations (with their semantics) provided. Liveness and bounded reuse
  do not have additional custom syntax and so supporting these would not require this step.
\end{enumerate}

\noindent
The list mirrors a list of steps that need to be done when supporting a new effectful computation
in a language that supports monadic do-notation. The step (3) corresponds to implementing a new
monad and (4) corresponds to adding monad-specific effectful operations. The step (2) applies when
using indexing to track effects more precisely \cite{effects-embedding}. The only step that does not
have effectful/monadic counterpart is (1).

When adding a new context-aware programming language support, much of the existing infrastructure
can be reused. This includes the implementation of the core coeffect and type checking rules and
also the translation for standard language constructs as well as the interpreter for the target
language.

% ==================================================================================================
%
%     #####
%    #     #   ##    ####  ######     ####  ##### #    # #####  # ######  ####
%    #        #  #  #      #         #        #   #    # #    # # #      #
%    #       #    #  ####  #####      ####    #   #    # #    # # #####   ####
%    #       ######      # #              #   #   #    # #    # # #           #
%    #     # #    # #    # #         #    #   #   #    # #    # # #      #    #
%     #####  #    #  ####  ######     ####    #    ####  #####  # ######  ####
%
% ==================================================================================================

@@

The prototype implementation illustrates a number of interesting aspects of coeffect systems.
Those appear as examples in the interactive essay (discussed in Section~\ref{sec:impl-essay}), but
we briefly review them in this section.

% --------------------------------------------------------------------------------------------------

\subsection{Typing context-aware programs}
\label{sec:impl-case-typing}

We first consider two case studies of how coeffect type checking works. The first one exposes the
resolution of the ambiguity in typing for implicit parameters and the second one exposes the
difference between flat and structural system for data-flow.

\paragraph{Abstraction for implicit parameters.}
As discussed in Section~\ref{sec:impl-theory-typing}, the implementation of the language with
implicit parameters resolves the ambiguity in the lambda abstraction by generating a coeffect
constraint that restricts the set of parameters required from the declaration-site to those that
are lexically available. Remaining parameters are required from the call-site. This is illustrated
by the following example:
%
\begin{equation*}
\begin{array}{l}
\kvd{let}~\ident{both}~=\\[-0.25em]
\quad \kvd{let}~\ident{?fst}~=~100~\kvd{in}\\[-0.25em]
\quad \kvd{fun}~\ident{trd}~\rightarrow~\ident{?fst}~+~\ident{?snd}~+~\ident{trd}~\kvd{in}\\[-0.25em]
\kvd{let}~\ident{?fst}~=~200~\kvd{in}\\[-0.25em]
\ident{both}~1
\end{array}
\end{equation*}
%
In this expression, the lambda function on line 3 requires implicit parameters \ident{?fst} and
\ident{?snd}. Since \ident{?fst} is available in scope, the type of \ident{both} is a function that
requires only \ident{?snd}. In the text-based notation used in the prototype, the type of the
function \ident{both} is: {\tt num -\{?snd:num\}-> num}.

\paragraph{Flat and structural data-flow.}
In flat data-flow, the context requirements of the body is required from both the declaration-site
and from the call-site. In structural data-flow, the context requirements are tracked separately
for each variable, which provides a more precise type. Consider the following two examples (the
\kvd{let} keyword is used to define a curried function of two arguments):
%
\begin{equation*}
\begin{array}{l}
\kvd{let}~\ident{oldy}~\ident{x}~\ident{y}~=~\ident{x}~+~\kvd{prev}~\ident{y}~\kvd{in}\\[-0.25em]
\ident{oldy}
\end{array}
\end{equation*}
%
When type checking the expression using the flat system, the type of \ident{oldy} is inferred as {\tt num -\{1\}-> num -\{1\}-> num},
but when using the structural system, the type becomes {\tt num -\{0\}-> num -\{1\}-> num}.

This illustrates the difference between the two - the flat system keeps only one annotation
for the whole body (which requires 1 past value). In lambda abstraction (or function declaration
written using \kvd{let}), this requirement is duplicated. The structural system keeps information
per-variable and so the resulting type reflects the fact that only the variable \idnet{y} appears
inside \kvd{prev}.

% --------------------------------------------------------------------------------------------------

@@

\begin{equation*}
\begin{array}{l}
let ?param = 10 in\\[-0.25em]
fun x -> ?param + ?other
\end{array}
\end{equation*}

\begin{equation*}
\begin{array}{l}
let (ctx2, ctx3) = split (duplicate finput) in\\[-0.25em]
let ctx1 = letimpl ?param (ctx2, 10) in\\[-0.25em]
fun x ->\\[-0.25em]
  let ctx4 = merge (x, ctx1) in\\[-0.25em]
  let (ctx5, ctx6) = split (duplicate ctx4) in\\[-0.25em]
  lookup ?param ctx5 + lookup ?other ctx6
\end{array}
\end{equation*}

\begin{equation*}
\begin{array}{l}
fun x -> (fun v -> prev v) (prev x)
\end{array}
\end{equation*}

\begin{equation*}
\begin{array}{l}
fun x ->\\[-0.25em]
  let ctx1 = merge (x, sinput) in\\[-0.25em]
  (fun v ->\\[-0.25em]
    let ctx3 = merge (v, choose_0 ctx1) in\\[-0.25em]
    counit (prev ctx3)) (cobind (fun ctx2 ->\\[-0.25em]
      counit (prev ctx2)) (choose_1 ctx1))
\end{array}
\end{equation*}
